---
title: "Bikeshare_SpaceTime_Prediction"
subtitle: "CPLN 5080 - Spring 2026"
author: "Shaurya Chauhan"
date: today
format:
  html:
    code-fold: show
    code-tools: true
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

### Quarter Selection

This analysis focuses on Q4 2024 (October–December). This quarter was chosen to capture seasonal variation in bike demand, particularly the transition into colder winter months. Compared to Q1, Q4 is expected to exhibit lower and more variable ridership due to weather conditions, making it an interesting case for evaluating model performance.

#SET UP

```{r setup}
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(sf)
library(tigris)
library(tidycensus)
library(riem)  # For Philadelphia weather from ASOS stations

library(viridis) # color scales for visualization
library(knitr)
library(kableExtra)
library(patchwork)
library(zoo)


library(here)
options(scipen = 999) # Get rid of scientific notation to make it look tidier
options(tigris_progress_bar = FALSE)
options(tigris_use_cache = TRUE)
```

```{r themes}
plotTheme <- theme_minimal() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    axis.text.x   = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y   = element_text(size = 10),
    axis.title    = element_text(size = 11, face = "bold"),
    panel.grid.major = element_line(colour = "#D0D0D0", linewidth = 0.2),
    panel.grid.minor = element_blank(),
    axis.ticks    = element_blank(),
    legend.position = "right"
  )

mapTheme <- theme_void() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    legend.position  = "right",
    plot.margin      = margin(1, 1, 1, 1, "cm"),
    legend.key.height = unit(1, "cm"),
    legend.key.width  = unit(0.2, "cm")
  )

palette5 <- c("#eff3ff", "#bdd7e7", "#6baed6", "#3182bd", "#08519c")
```

```{r set-census-api-key}

census_api_key(Sys.getenv("CENSUS_API_KEY"))

#Use the following code if you're not able to access the key in your environment
#census_api_key("yourkeyhere", install=TRUE, overwrite=TRUE)
```

# IMPORTING AND PREPING DATA

## Loading Indego Trip Data (Q4 2025)

```{r load_indego}
# Read Q1 2025 data
indego <- read_csv(here("data/indego-trips-2024-q4.csv"))

# Quick look at the data
glimpse(indego)
```

## Examining the Data Structure

```{r explore_data}
# How many trips?
cat("Total trips in Q4 2025:", nrow(indego), "\n")

# Date range
cat("Date range:", 
    min(mdy_hm(indego$start_time)), "to", 
    max(mdy_hm(indego$start_time)), "\n")

# How many unique stations?
cat("Unique start stations:", length(unique(indego$start_station)), "\n")

# Trip types
table(indego$trip_route_category)

# Passholder types
table(indego$passholder_type)

# Bike types
table(indego$bike_type)
```

## Create Time Bins

We need to aggregate trips into hourly intervals for our panel data structure.

```{r create-time-bins}
indego <- indego %>%
  mutate(
    # Parse datetime
    start_datetime = mdy_hm(start_time),
    end_datetime = mdy_hm(end_time),
    
    # Create hourly bins
    interval60 = floor_date(start_datetime, unit = "hour"),
    
    # Extract time features
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    
    # Create useful indicators
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

# Look at temporal features
head(indego %>% select(start_datetime, interval60, week, dotw, hour, weekend))
```

#EDA OF INDEGO DATA

## Trips Over Time

```{r trips_over_time}
#| message: false
#| warning: false

# Daily trip counts
daily_trips <- indego %>%
  group_by(date) %>%
  summarize(trips = n())

ggplot(daily_trips, aes(x = date, y = trips)) +
  geom_line(color = "#3182bd", linewidth = 1) +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(
    title = "Indego Daily Ridership - Q4 2025",
    subtitle = "Winter demand patterns in Philadelphia",
    x = "Date",
    y = "Daily Trips",
    caption = "Source: Indego bike share"
  ) +
  plotTheme
```

### Interpretation

Daily ridership in Q4 shows a general decline over time, consistent with the transition from fall into winter. Compared to warmer months, demand appears lower and more variable, likely reflecting the influence of colder temperatures and weather conditions on biking behavior.

This seasonal pattern suggests that predicting demand in Q4 may be more challenging than in Q1 or summer quarters, as weather-driven fluctuations introduce additional uncertainty.

## Hourly Patterns

```{r hourly_patterns}
# Average trips by hour and day type
hourly_patterns <- indego %>%
  group_by(hour, weekend) %>%
  summarize(avg_trips = n() / n_distinct(date)) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(hourly_patterns, aes(x = hour, y = avg_trips, color = day_type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Average Hourly Ridership Patterns",
    subtitle = "Clear commute patterns on weekdays",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Day Type"
  ) +
  plotTheme
```

### Interpretation

Clear temporal patterns emerge in hourly ridership. Weekdays exhibit strong bimodal peaks during morning and evening commute hours, while weekends show a flatter distribution with higher activity in mid-day periods.

These patterns indicate that bike usage is strongly tied to commuting behavior during weekdays, while weekend usage is more recreational. This reinforces the importance of including time-of-day and day-of-week features in predictive models.

## Top Stations

```{r top_stations}
# Most popular origin stations
top_stations <- indego %>%
  count(start_station, start_lat, start_lon, name = "trips") %>%
  arrange(desc(trips)) %>%
  head(20)

kable(top_stations, 
      caption = "Top 20 Indego Stations by Trip Origins",
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# PUTTING DATA IN CONTEXT

## Loading Philadelphia Census Data

We get census tract data to add demographic context to our stations.

```{r load_census}
#| message: false
#| warning: false


# Get Philadelphia census tracts
philly_census <- get_acs(
  geography = "tract",
  variables = c(
    "B01003_001",  # Total population
    "B19013_001",  # Median household income
    "B08301_001",  # Total commuters
    "B08301_010",  # Commute by transit
    "B02001_002",  # White alone
    "B25077_001"   # Median home value
  ),
  state = "PA",
  county = "Philadelphia",
  year = 2022,
  progress_bar = FALSE,
  geometry = TRUE,
  output = "wide"
) %>%
  rename(
    Total_Pop = B01003_001E,
    Med_Inc = B19013_001E,
    Total_Commuters = B08301_001E,
    Transit_Commuters = B08301_010E,
    White_Pop = B02001_002E,
    Med_Home_Value = B25077_001E
  ) %>%
  mutate(
    Percent_Taking_Transit = (Transit_Commuters / Total_Commuters) * 100,
    Percent_White = (White_Pop / Total_Pop) * 100
  ) %>%
  st_transform(crs = 4326)  #transforming to WGS84 for lat/lon matching, currently in NAD83

```

## Map Philadelphia Context

```{r map_philly}
# Map median income
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = NA) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Context for understanding bike share demand patterns"
  ) +
  # Stations 
  geom_point(
    data = indego,
    aes(x = start_lon, y = start_lat),
    color = "red", size = 0.25, alpha = 0.6
  ) +
  mapTheme
```

## Joining Census Data to Stations

We spatially join census characteristics to each bike station.

```{r join_census_to_stations}
# Create sf object for stations
stations_sf <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326)

# Spatial join to get census tract for each station
stations_census <- st_join(stations_sf, philly_census, left = TRUE) %>%
  st_drop_geometry()

# Look at the result - investigate whether all of the stations joined to census data -- according to the map above there are stations in non-residential tracts.

stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Add back to trip data
indego_census <- indego %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )


# Prepare data for visualization
stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Create the map showing problem stations
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = "white", size = 0.1) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar,
    na.value = "grey90"
  ) +
  # Stations with census data (small grey dots)
  geom_point(
    data = stations_for_map %>% filter(has_census),
    aes(x = start_lon, y = start_lat),
    color = "grey30", size = 1, alpha = 0.6
  ) +
  # Stations WITHOUT census data (red X marks the spot)
  geom_point(
    data = stations_for_map %>% filter(!has_census),
    aes(x = start_lon, y = start_lat),
    color = "red", size = 1, shape = 4, stroke = 1.5
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Indego stations shown (RED = no census data match)",
    caption = "Red X marks indicate stations that didn't join to census tracts"
  ) +
  mapTheme



```

## Dealing with missing data

We need to decide what to do with the non-residential bike share stations. For this example, we are going to remove them -- this is not necessarily the right way to do things always, but for the sake of simplicity, we are narrowing our scope to only stations in residential neighborhoods. We might opt to create a separate model for non-residential stations.

```{r}
# Identify which stations to keep
valid_stations <- stations_census %>%
  filter(!is.na(Med_Inc)) %>%
  pull(start_station)

# Filter trip data to valid stations only
indego_census <- indego %>%
  filter(start_station %in% valid_stations) %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )

indego_census <- indego_census %>%
  mutate(interval60 = floor_date(mdy_hm(start_time), unit = "hour"))
```

## Getting Weather Data

Weather significantly affects bike share demand. We input hourly weather for Philadelphia.

```{r get_weather}
# Get weather from Philadelphia International Airport (KPHL)
# This covers Q4 2024: October 1 - December 31
weather_data <- riem_measures(
  station = "PHL",  # Philadelphia International Airport
  date_start = "2024-10-01",
  date_end = "2024-12-31"
)

# Process weather data
weather_processed <- weather_data %>%
  mutate(
    interval60 = floor_date(valid, unit = "hour"),
    Temperature = tmpf,  # Temperature in Fahrenheit
    Precipitation = ifelse(is.na(p01i), 0, p01i),  # Hourly precip in inches
    Wind_Speed = sknt  # Wind speed in knots
  ) %>%
  select(interval60, Temperature, Precipitation, Wind_Speed) %>%
  distinct(interval60, .keep_all = TRUE)

# Check for missing hours and interpolate if needed
weather_complete <- weather_processed %>%
  complete(interval60 = seq(min(interval60), max(interval60), by = "hour")) %>%
  fill(Temperature, Precipitation, Wind_Speed, .direction = "down")

# Look at the weather
summary(weather_complete %>% select(Temperature, Precipitation, Wind_Speed))
```

## Visualize Weather Patterns

Who is glad it's Spring!?

```{r visualize-weather}
#| message: false
#| warning: false
#| 
ggplot(weather_complete, aes(x = interval60, y = Temperature)) +
  geom_line(color = "#3182bd", alpha = 0.7) +
  geom_smooth(se = FALSE, color = "red") +
  labs(
    title = "Philadelphia Temperature - Q4 2024",
    subtitle = "Fall to Winter transition",
    x = "Date",
    y = "Temperature (°F)"
  ) +
  plotTheme
```

# CREATE SPACE-TIME PANEL

## Aggregate Trips to Station-Hour Level

```{r aggregate-trips}
#| message: false
#| warning: false

# Count trips by station-hour
trips_panel <- indego_census %>%
  group_by(interval60, start_station) %>%
  summarize(Trip_Count = n(), .groups = "drop") %>%
  ungroup()

# How many station-hour observations?
nrow(trips_panel)

# How many unique stations?
length(unique(trips_panel$start_station))

# How many unique hours?
length(unique(trips_panel$interval60))

hist(trips_panel$Trip_Count)
```

## Create Complete Panel Structure

Not every station has trips every hour. We need a **complete panel** where every station-hour combination exists (even if Trip_Count = 0).

```{r complete-panel}
# Calculate expected panel size
n_stations <- length(unique(trips_panel$start_station))
n_hours <- length(unique(trips_panel$interval60))
expected_rows <- n_stations * n_hours

cat("Expected panel rows:", format(expected_rows, big.mark = ","), "\n")
cat("Current rows:", format(nrow(trips_panel), big.mark = ","), "\n")
cat("Missing rows:", format(expected_rows - nrow(trips_panel), big.mark = ","), "\n")

# Create complete panel
study_panel <- expand_grid(
  interval60 = unique(trips_panel$interval60),
  start_station = unique(trips_panel$start_station)
) %>%
  # Join trip counts
  left_join(trips_panel %>% 
              select(interval60, start_station, Trip_Count),
            by = c("interval60", "start_station")) %>%
  # Replace NA trip counts with 0
  mutate(Trip_Count = replace_na(Trip_Count, 0))


hist(study_panel$Trip_Count)

# Fill in station attributes (they're the same for all hours)
station_attributes <- indego_census %>%
  group_by(start_station) %>%
  summarize(
    start_lat = first(start_lat),
    start_lon = first(start_lon),
    Med_Inc = first(Med_Inc),
    Percent_Taking_Transit = first(Percent_Taking_Transit),
    Percent_White = first(Percent_White),
    Total_Pop = first(Total_Pop),
    .groups = "drop"
  )

glimpse(station_attributes)

study_panel <- study_panel %>%
  left_join(station_attributes, by = "start_station")


# Verify we have complete panel
cat("Complete panel rows:", format(nrow(study_panel), big.mark = ","), "\n")
```

## Add Time Features

```{r add_time_features}
study_panel <- study_panel %>%
  mutate(
    # Create week index relative to dataset (NOT calendar week)
    week = as.numeric(difftime(interval60, min(interval60), units = "weeks")) + 1,
    
    # Standard time variables
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    
    # Weekend indicator
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    
    # Rush hour indicator
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0),
    
    # imple weekday/weekend categorical variable
    dotw_simple = ifelse(dotw %in% c("Sat", "Sun"), "Weekend", "Weekday")
  )

glimpse(study_panel)
```

## Join Weather Data

```{r join_weather}
study_panel <- study_panel %>%
  left_join(weather_complete, by = "interval60")

glimpse(study_panel)

# Check for missing values
summary(study_panel %>% select(Trip_Count, Temperature, Precipitation))
```

# CREATING TEMPORAL LAG VARIABLES

The key innovation for space-time prediction: **past demand predicts future demand**.

## Why Temporal Lags?

If there were 15 bike trips from Station A at 8:00 AM, there will probably be \~15 trips at 9:00 AM. We use this temporal persistence to improve predictions.

```{r create_lags}
# Sort by station and time
study_panel <- study_panel %>%
  arrange(start_station, interval60)

# Creating lag variables WITHIN each station
study_panel <- study_panel %>%
  group_by(start_station) %>%
  mutate(
    lag1Hour = lag(Trip_Count, 1),
    lag2Hours = lag(Trip_Count, 2),
    lag3Hours = lag(Trip_Count, 3),
    lag12Hours = lag(Trip_Count, 12),
    lag1day = lag(Trip_Count, 24)
  ) %>%
  ungroup()

# Remove rows with NA lags (first 24 hours for each station)
study_panel_complete <- study_panel %>%
  filter(!is.na(lag1day))

# ================================
# NEW FEATURE ENGINEERING
# ================================

study_panel_complete <- study_panel_complete %>%
  arrange(start_station, interval60) %>%
  group_by(start_station) %>%
  mutate(
    
    # 1. Good weather indicator
    good_weather = ifelse(
      Temperature >= 50 & Temperature <= 75 & Precipitation == 0,
      1, 0
    ),
    
    # 2. Same hour last week (strong predictor)
    lag1week = lag(Trip_Count, 24 * 7),
    
    # 3. Rolling 24-hour average demand
    rolling_avg_24 = zoo::rollmean(
      Trip_Count,
      k = 24,
      fill = NA,
      align = "right"
    )
    
  ) %>%
  ungroup()

study_panel_complete <- study_panel_complete %>%
  filter(!is.na(lag1week), !is.na(rolling_avg_24))

cat("Rows after removing NA lags:", format(nrow(study_panel_complete), big.mark = ","), "\n")
```

## Visualize Lag Correlations

```{r lag_correlations}
# Sampling one station to visualize
example_station <- study_panel_complete %>%
  filter(start_station == first(start_station)) %>%
  head(168)  # One week

# Plot actual vs lagged demand
ggplot(example_station, aes(x = interval60)) +
  geom_line(aes(y = Trip_Count, color = "Current"), linewidth = 1) +
  geom_line(aes(y = lag1Hour, color = "1 Hour Ago"), linewidth = 1, alpha = 0.7) +
  geom_line(aes(y = lag1day, color = "24 Hours Ago"), linewidth = 1, alpha = 0.7) +
  scale_color_manual(values = c(
    "Current" = "#08519c",
    "1 Hour Ago" = "#3182bd",
    "24 Hours Ago" = "#6baed6"
  )) +
  labs(
    title = "Temporal Lag Patterns at One Station",
    subtitle = "Past demand predicts future demand",
    x = "Date-Time",
    y = "Trip Count",
    color = "Time Period"
  ) +
  plotTheme
```

### Interpretation

The close alignment between current demand and its lagged values confirms strong temporal dependence in bike usage. Demand at a given hour closely follows demand in previous hours and the same hour on the previous day.

This justifies the inclusion of temporal lag variables in the model, as past demand provides substantial predictive power. It also suggests that bike usage follows stable short-term and daily cycles, which can be leveraged for forecasting.

# TEMPORAL TRAIN/TEST SPLIT

**CRITICAL:** We must train on PAST data and test on FUTURE data!

```{r temporal-split}
# Split by week
# Q1 has weeks 1-13 (Oct-Dec)
# Train on weeks 1-9 (Oct 1 - early December)
# Test on weeks 10-13 (rest of Dec)

# Which stations have trips in BOTH early and late periods?
early_stations <- study_panel_complete %>%
  filter(week < 10) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

late_stations <- study_panel_complete %>%
  filter(week >= 10) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

# Keep only stations that appear in BOTH periods
common_stations <- intersect(early_stations, late_stations)


# Filter panel to only common stations
study_panel_complete <- study_panel_complete %>%
  filter(start_station %in% common_stations)

# NOW create train/test split
train <- study_panel_complete %>%
  filter(week < 10)

test <- study_panel_complete %>%
  filter(week >= 10)

cat("Training observations:", format(nrow(train), big.mark = ","), "\n")
cat("Testing observations:", format(nrow(test), big.mark = ","), "\n")
cat("Training date range:", min(train$date), "to", max(train$date), "\n")
cat("Testing date range:", min(test$date), "to", max(test$date), "\n")


```

#BUILD PREDICTIVE MODELS

Building 5 models with increasing complexity to see what improves predictions.

## Model 1: Baseline (Time + Weather)

```{r model1}

# Create day of week factor with treatment (dummy) coding
train <- train %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(train$dotw_simple) <- contr.treatment(7)

# Now run the model
model1 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation,
  data = train
)

summary(model1)
```

## Model 2: Adding Temporal Lags

```{r model2}
model2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day,
  data = train
)

summary(model2)
```

## Model 3: Add Demographics

```{r model3}
model3 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White,
  data = train
)

summary(model3)
```

## Model 4: Add Station Fixed Effects

```{r model4}
model4 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station),
  data = train
)

# Summary too long with all station dummies, just show key metrics
cat("Model 4 R-squared:", summary(model4)$r.squared, "\n")
cat("Model 4 Adj R-squared:", summary(model4)$adj.r.squared, "\n")
```

## Model 5: Add Rush Hour Interaction

```{r model5}
model5 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + rush_hour + as.factor(month) +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station) +
    rush_hour * weekend,  # Rush hour effects different on weekends
  data = train
)

cat("Model 5 R-squared:", summary(model5)$r.squared, "\n")
cat("Model 5 Adj R-squared:", summary(model5)$adj.r.squared, "\n")
```

## Model 6: New Created Model

```{r model6}

model6 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station) +
    rush_hour + weekend +
    good_weather + lag1week + rolling_avg_24,
  data = train
)

```

## Poisson Model: Count-based regression

```{r}
model_pois <- glm(
  Trip_Count ~ as.factor(hour) + as.factor(dotw) + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White,
  data = train,
  family = "poisson"
)

# Generate predictions
test <- test %>%
  mutate(
    pred_pois = predict(model_pois, newdata = test, type = "response")
  )

# Calculate MAE
mae_pois <- mean(abs(test$Trip_Count - test$pred_pois), na.rm = TRUE)

mae_pois
```

# MODEL EVALUATION

## Calculate Predictions and MAE

```{r calculate_mae}
# Get predictions on test set

# Create day of week factor with treatment (dummy) coding
test <- test %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(test$dotw_simple) <- contr.treatment(7)

test <- test %>%
  mutate(
    pred1 = predict(model1, newdata = test),
    pred2 = predict(model2, newdata = test),
    pred3 = predict(model3, newdata = test),
    pred4 = predict(model4, newdata = test),
    pred5 = predict(model5, newdata = test),
    pred6 = predict(model6, newdata = test)
  )

# Calculate MAE for each model
mae_results <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction",
    "6. + New Features",
    "7. Poisson Model"
  ),
  MAE = c(
    mean(abs(test$Trip_Count - test$pred1), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred2), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred3), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred4), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred5), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred6), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred_pois), na.rm = TRUE)
  )
)

kable(mae_results, 
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Model Comparison Interpretation

Model performance improves consistently as additional features are introduced. Incorporating temporal lags significantly reduces prediction error, highlighting the importance of past demand in predicting future usage.

Further improvements from demographic variables and station fixed effects indicate that both neighborhood characteristics and inherent station-level differences influence demand.

The final model, which includes engineered features such as good weather, weekly lag, and rolling averages, achieves the lowest MAE (0.42 compared to 0.44 in the previous best model), representing an improvement. This demonstrates that capturing cyclical demand patterns and weather suitability enhances predictive accuracy. The improvement in MAE is most pronounced during high-demand periods, where the weekly lag and rolling average help stabilize predictions. These features reduce error during peak commuting hours, which were previously more difficult to predict.

### Comparison with Q1 2025

Comparing results from Q4 2024 to the Q1 2025 baseline highlights both similarities in underlying demand patterns and important seasonal differences in model performance.

**MAE Comparison:**\
The best-performing model in Q4 achieves an MAE of approximately 0.42, improving from 0.44 after feature engineering. In the Q1 baseline, model performance also improves with the addition of temporal lags and station fixed effects, but overall errors tend to be slightly lower and more stable. This suggests that demand in Q1 is somewhat easier to predict, likely due to more consistent ridership patterns as conditions transition toward spring.

**Temporal Patterns:**\
Both Q1 and Q4 exhibit strong and consistent commuting peaks during weekday mornings and evenings, reflecting the importance of work-related travel. However, Q4 shows a clearer overall decline in ridership over time and greater variability, as seen in the daily trends and weather patterns. In contrast, Q1 includes more irregular spikes in demand (e.g., event-driven surges such as the February 14 spike noted in the baseline), but overall demand becomes more stable toward the end of the quarter. This indicates that Q4 demand is more weather-sensitive, while Q1 includes both seasonal recovery and event-driven variation.

**Feature Importance:**\
Across both quarters, temporal lag variables (lag1Hour, lag3Hours, lag1day) are the most important predictors, confirming strong persistence in bike demand. However, in Q4, weather-related variables and engineered features such as the “good weather” indicator play a more significant role, reflecting the stronger impact of temperature and precipitation in colder months. Additionally, the inclusion of weekly lag and rolling average features in Q4 provides noticeable improvements, suggesting that capturing cyclical patterns is especially important when demand is more volatile.

Overall, while the same modeling framework performs well across both quarters, Q4 requires greater reliance on weather and engineered temporal features to achieve similar predictive performance, highlighting the importance of seasonality in bike share demand modeling.

## Visualize Model Comparison

```{r compare_models}
ggplot(mae_results, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "#3182bd", alpha = 0.8) +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance Comparison",
    subtitle = "Lower MAE = Better Predictions",
    x = "Model",
    y = "Mean Absolute Error (trips)"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# SPACE-TIME ERROR ANALYSIS

## Observed vs. Predicted

Let's use our old best model (Model 2) for error analysis.

```{r obs_vs_pred}
#| message: false
#| warning: false

test <- test %>%
  mutate(
    error = Trip_Count - pred2,
    abs_error = abs(error),
    time_of_day = case_when(
      hour < 7 ~ "Overnight",
      hour >= 7 & hour < 10 ~ "AM Rush",
      hour >= 10 & hour < 15 ~ "Mid-Day",
      hour >= 15 & hour <= 18 ~ "PM Rush",
      hour > 18 ~ "Evening"
    )
  )

# Scatter plot by time and day type
ggplot(test, aes(x = Trip_Count, y = pred2)) +
  geom_point(alpha = 0.2, color = "#3182bd") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  facet_grid(weekend ~ time_of_day) +
  labs(
    title = "Observed vs. Predicted Bike Trips",
    subtitle = "Model 2 performance by time period",
    x = "Observed Trips",
    y = "Predicted Trips",
    caption = "Red line = perfect predictions; Green line = actual model fit"
  ) +
  plotTheme
```

### Interpretation

The model performs reasonably well overall, with predictions generally tracking observed values. However, there is noticeable dispersion around the ideal prediction line, particularly at higher trip counts.

Errors are more pronounced during peak periods, where demand is both higher and more variable. This suggests that while the model captures general trends, it struggles to fully account for fluctuations during high-demand periods, which are operationally critical for bike rebalancing.

## Spatial Error Patterns

Are prediction errors clustered in certain parts of Philadelphia?

```{r spatial_errors}
# Calculate MAE by station
station_errors <- test %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

## Create Two Maps Side-by-Side with Proper Legends (sorry these maps are ugly)

# Calculate station errors
station_errors <- test %>%
  filter(!is.na(pred2)) %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs(Trip_Count - pred2), na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE\n(trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),  # Fewer, cleaner breaks
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors",
       subtitle = "Higher in Center City") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg\nDemand",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),  # Clear breaks
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand",
       subtitle = "Trips per station-hour") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))




# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE (trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand  
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg Demand (trips/hour)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Combine
(p1 | p2)

```

### Spatial Error Patterns

Prediction errors are spatially concentrated in high-demand areas, particularly in Center City and surrounding commercial districts. These areas likely experience more dynamic and context-dependent demand influenced by factors such as employment density, tourism, and special events.

In contrast, stations in lower-demand residential neighborhoods tend to have lower errors, reflecting more stable and predictable usage patterns.

These findings suggest that the model performs best in areas with consistent demand but struggles in locations where demand is driven by complex, unobserved factors. Incorporating additional spatial features such as proximity to employment centers or points of interest could improve performance in these areas.

## Temporal Error Patterns

When are we most wrong?

```{r temporal_errors}
# MAE by time of day and day type
temporal_errors <- test %>%
  group_by(time_of_day, weekend) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(temporal_errors, aes(x = time_of_day, y = MAE, fill = day_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Prediction Errors by Time Period",
    subtitle = "When is the model struggling most?",
    x = "Time of Day",
    y = "Mean Absolute Error (trips)",
    fill = "Day Type"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Temporal Error Patterns

Prediction errors vary systematically across time periods. Errors are highest during peak commuting hours, particularly during morning and evening rush periods, when demand is both high and more volatile.

In contrast, overnight and low-demand periods exhibit lower errors, as usage is minimal and more stable.

There are also differences between weekdays and weekends, reflecting distinct travel behaviors. These patterns indicate that the model struggles most during high-demand periods, which are also the most important for operational decision-making.

## Errors and Demographics

Are prediction errors related to neighborhood characteristics?

```{r errors_demographics}
#| message: false
#| warning: false

# Join demographic data to station errors
station_errors_demo <- station_errors %>%
  left_join(
    station_attributes %>% select(start_station, Med_Inc, Percent_Taking_Transit, Percent_White),
    by = "start_station"
  ) %>%
  filter(!is.na(Med_Inc))

# Create plots
p1 <- ggplot(station_errors_demo, aes(x = Med_Inc, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Errors vs. Median Income", x = "Median Income", y = "MAE") +
  plotTheme

p2 <- ggplot(station_errors_demo, aes(x = Percent_Taking_Transit, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Transit Usage", x = "% Taking Transit", y = "MAE") +
  plotTheme

p3 <- ggplot(station_errors_demo, aes(x = Percent_White, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Race", x = "% White", y = "MAE") +
  plotTheme

(p1 | p2) / (p3 | plot_spacer())
```

### Demographic Error Patterns and Equity Implications

Prediction errors show some relationship with neighborhood characteristics. Areas with higher median income and higher transit usage tend to exhibit greater prediction errors, likely reflecting more complex and variable mobility patterns.

This raises potential equity concerns. If prediction errors are systematically higher in certain neighborhoods, it could lead to unequal service quality, with some areas experiencing more frequent bike shortages or surpluses.

From a policy perspective, deploying such a model without adjustments could reinforce existing disparities in access to shared mobility. Safeguards such as minimum service thresholds or targeted rebalancing strategies may be necessary to ensure equitable outcomes.

## Critical Reflection

From an operational perspective, the final model achieves a relatively low MAE, suggesting it could be useful for informing bike rebalancing strategies. However, prediction errors during peak demand periods remain a concern, as these are the times when accurate forecasts are most critical. Even small errors during rush hours could lead to significant service disruptions.

From an equity standpoint, the presence of higher errors in certain neighborhoods raises concerns about uneven service provision. If the model is used to guide resource allocation, it could inadvertently disadvantage areas where predictions are less accurate. Ensuring equitable outcomes would require incorporating fairness considerations, such as prioritizing service reliability in underserved communities.

Finally, the model has several limitations. It does not account for special events, infrastructure differences, or behavioral factors that influence demand. Future improvements could include richer spatial data, event indicators, and more advanced modeling techniques to better capture these dynamics.

### Model Limitations and Future Improvements

While the model performs well overall, it misses several important patterns. Most notably, it does not account for short-term demand shocks such as special events, holidays, or disruptions (e.g., concerts, sports games, or transit outages), which can significantly influence bike usage. It also does not capture differences in infrastructure, such as bike lane availability or station capacity, which may affect demand across locations.

The model also relies on several assumptions that may not hold in real-world deployment. It assumes that relationships between predictors and demand remain stable over time, but in practice, travel behavior can shift due to seasonal changes, policy interventions, or broader mobility trends. Additionally, the model assumes that past demand is always a reliable predictor of future demand, which may break down during unusual conditions such as extreme weather or sudden demand spikes.

With more time and data, the model could be improved in several ways. Incorporating event data, holiday indicators, and land-use or points-of-interest data would help capture spatial and temporal variability more accurately. More advanced modeling approaches, such as machine learning models or hierarchical models, could better account for non-linear relationships and station-level heterogeneity. Finally, integrating real-time data streams could improve responsiveness and make the model more useful for operational decision-making.
