---
title: "Lab 4: Spatial Predictive Analysis"
subtitle: "Analyzing 311 Violations"
author: "Shaurya Chauhan"
date: today
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---



## Overview

This analysis builds a spatial predictive model of burglary incidents in Chicago using neighborhood-level indicators of urban disorder. 

The primary predictor used in this study is 311 garbage cart complaints, which serve as a proxy for neighborhood maintenance and disorder. Prior research suggests that visible signs of neglect may be associated with higher crime risk by reducing informal social control.

The objective is to evaluate whether incorporating spatial features derived from garbage complaints improves the prediction of burglary incidents compared to simple spatial smoothing methods such as kernel density estimation (KDE).


```{r setup}
#| message: false
#| warning: false

# Load required packages
library(tidyverse)
library(sf)
library(here)
library(viridis)
library(terra)
library(spdep)
library(FNN)
library(MASS)
library(patchwork)
library(knitr)
library(kableExtra)
library(classInt)

library(spatstat.geom)
library(spatstat.explore)

options(scipen = 999)
set.seed(5080)

theme_crime <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 1),
      plot.subtitle = element_text(color = "gray30", size = base_size - 1),
      legend.position = "right",
      panel.grid.minor = element_blank(),
      axis.text = element_blank(),
      axis.title = element_blank()
    )
}

theme_set(theme_crime())

cat("✓ All packages loaded successfully!\n")
cat("✓ Working directory:", getwd(), "\n")
```


## Exercise 1.1: Load Chicago Spatial Data

```{r load-boundaries}
#| message: false

# Load police districts (used for spatial cross-validation)
policeDistricts <- 
  st_read("https://data.cityofchicago.org/api/geospatial/24zt-jpfn?method=export&format=GeoJSON") %>%
  st_transform('ESRI:102271') %>%
  dplyr::select(District = dist_num)

# Load police beats (smaller administrative units)
policeBeats <- 
  st_read("https://data.cityofchicago.org/api/geospatial/n9it-hstw?method=export&format=GeoJSON") %>%
  st_transform('ESRI:102271') %>%
  dplyr::select(Beat = beat_num)

# Load Chicago boundary
chicagoBoundary <- 
  st_read("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/DATA/Chapter5/chicagoBoundary.geojson") %>%
  st_transform('ESRI:102271')

cat("✓ Loaded spatial boundaries\n")
cat("  - Police districts:", nrow(policeDistricts), "\n")
cat("  - Police beats:", nrow(policeBeats), "\n")
```

The analysis begins by loading key spatial boundary datasets for Chicago, 
including police districts, police beats, and the city boundary. These 
geographies provide the spatial framework for the analysis.

Police districts are particularly important because they are used for spatial 
cross-validation, allowing the model to be evaluated on unseen geographic areas.

All datasets are transformed to a common coordinate reference system 
(ESRI:102271) to ensure consistency in spatial operations.


## Exercise 1.2: Load Burglary Data

```{r load-burglaries}
#| message: false

burglaries <- st_read(here("data", "burglaries.shp")) %>% 
  st_transform('ESRI:102271')

glimpse(burglaries)

cat("\n✓ Loaded burglary data\n")
cat("  - Number of burglaries:", nrow(burglaries), "\n")
cat("  - CRS:", st_crs(burglaries)$input, "\n")
```

The dataset contains burglary incidents across Chicago for 2017. The data is 
stored as a spatial point dataset and transformed into the ESRI:102271 
coordinate system to enable accurate distance-based analysis.


## Exercise 1.3: Visualize Crime Patterns

```{r visualize-burglaries}
#| fig-width: 10
#| fig-height: 5

p1 <- ggplot() + 
  geom_sf(data = chicagoBoundary, fill = "gray95", color = "gray60") +
  geom_sf(data = burglaries, color = "#d62828", size = 0.1, alpha = 0.4) +
  labs(
    title = "Burglary Locations",
    subtitle = paste0("Chicago 2017, n = ", nrow(burglaries))
  )

p2 <- ggplot() + 
  geom_sf(data = chicagoBoundary, fill = "gray95", color = "gray60") +
  geom_density_2d_filled(
    data = data.frame(st_coordinates(burglaries)),
    aes(X, Y),
    alpha = 0.7,
    bins = 8
  ) +
  scale_fill_viridis_d(option = "plasma", direction = -1) +
  labs(
    title = "Burglary Density Surface",
    subtitle = "Kernel density estimation"
  )

p1 + p2
```
Burglary incidents are spatially clustered rather than randomly distributed 
across Chicago- especially around the south-eastern and central Chicago areas. The density map reveals concentrated hotspots of crime, 
indicating that these neighborhoods experience significantly higher levels 
of burglary.

This clustering suggests that spatial features and neighborhood conditions 
play a critical role in explaining crime patterns, making spatial modeling 
essential for prediction.


## Exercise 2.1: Create Spatial Analysis Grid

```{r create-fishnet}

fishnet <- st_make_grid(
  chicagoBoundary,
  cellsize = 1000,
  square = TRUE
) %>%
  st_as_sf() %>%
  st_intersection(chicagoBoundary)

fishnet$grid_id <- 1:nrow(fishnet)

nrow(fishnet)
```

A regular grid (fishnet) is created to aggregate crime data into spatial units. 
Each grid cell represents a 1000-foot square area, allowing crime counts and 
other features to be measured consistently across space.

This approach converts point-level data into area-based observations, which 
are required for regression modeling.


## Exercise 2.2: Aggregate Crime Counts

```{r crime-counts}

fishnet$crime_count <- lengths(
  st_intersects(fishnet, burglaries)
)

summary(fishnet$crime_count)
```

Crime counts are calculated for each grid cell by counting the number of 
burglary incidents that fall within each spatial unit. This creates the 
dependent variable for the regression model.

The distribution of crime counts is typically skewed, with many cells 
having low counts and a smaller number of cells experiencing high crime.


## Exercise 3.1: Load Garbage Cart Data

```{r load-garbage}
#| message: false

garbage_raw <- read_csv(here("data", "Chicago_GarbageCarts.csv"))

# Remove rows with missing coordinates
garbage_clean <- garbage_raw %>%
  drop_na(Longitude, Latitude)

# Convert to sf
garbage <- garbage_clean %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform('ESRI:102271')

cat("✓ Garbage data loaded\n")
cat("  - Original rows:", nrow(garbage_raw), "\n")
cat("  - After cleaning:", nrow(garbage), "\n")
```

Garbage cart locations are incorporated as an additional spatial feature. 
The dataset is converted into a spatial format and projected into the same 
coordinate system as the rest of the analysis.

This variable serves as a proxy for neighborhood conditions and municipal 
service patterns, which may be associated with crime levels.


## Exercise 3.2: Aggregate Garbage Carts to Grid

```{r garbage-counts}

fishnet$garbage_count <- lengths(
  st_intersects(fishnet, garbage)
)

summary(fishnet$garbage_count)
```

Garbage cart counts are calculated for each grid cell to capture the spatial 
distribution of waste management infrastructure.

Areas with higher concentrations of garbage carts may reflect higher population 
density or different neighborhood characteristics, which could influence crime patterns.


## Exercise 3.3: Create Spatial Features

```{r spatial-features}

# Get coordinates of grid centroids
coords <- st_coordinates(st_centroid(fishnet))

# Use knearneigh (spdep-compatible)
knn_obj <- knearneigh(coords, k = 4)

# Convert to neighbors list
neighbors <- knn2nb(knn_obj)

# Create spatial weights
weights <- nb2listw(neighbors, style = "W")

# Spatial lag of crime
fishnet$crime_lag <- lag.listw(weights, fishnet$crime_count)

summary(fishnet$crime_lag)
```

Spatial features are created to capture spatial dependence in crime patterns. 
Nearest neighbor distances measure how close each grid cell is to others, while 
spatial lag captures the average crime levels in neighboring cells.

These features allow the model to account for spatial autocorrelation, where 
crime in one area is influenced by crime in nearby areas.


## Exercise 3.4: Attach Police Districts

```{r attach-districts}

fishnet <- st_join(fishnet, policeDistricts)

table(fishnet$District)
```

Police district identifiers are attached to each grid cell. These are later 
used for spatial cross-validation, ensuring that model evaluation is performed 
across distinct geographic areas rather than random splits.


## Exercise 4.1: Negative Binomial Model

```{r nb-model}

nb_model <- glm.nb(
  crime_count ~ garbage_count + crime_lag,
  data = fishnet
)

summary(nb_model)
```


A Negative Binomial regression model is used to predict burglary counts. This 
model is appropriate because crime data are count-based and typically exhibit 
overdispersion, where the variance exceeds the mean.

The model includes garbage cart density and spatial lag of crime. Garbage carts 
serve as a proxy for neighborhood conditions and service patterns, while the 
spatial lag captures the influence of nearby areas. Together, these variables 
allow the model to account for both environmental and spatial drivers of crime.

The Negative Binomial regression results indicate that both garbage cart density 
and spatial lag are statistically significant predictors of burglary counts.

The coefficient for **garbage_count (0.00497, p < 0.001)** is positive and highly 
significant, suggesting that areas with more garbage carts tend to experience 
slightly higher burglary counts. While the magnitude is modest, this variable 
likely captures underlying neighborhood characteristics such as density or 
activity levels rather than a direct causal relationship.

The coefficient for **crime_lag (0.0756, p < 0.001)** is substantially larger and 
also highly significant. This indicates strong spatial dependence, meaning that 
crime in a given location is heavily influenced by crime levels in neighboring 
areas. This is consistent with well-established patterns of crime clustering in 
urban environments.



## Exercise 4.2: Model Predictions

```{r predictions}

fishnet$predicted <- predict(nb_model, type = "response")

summary(fishnet$predicted)
```



## Exercise 4.3: Predicted Crime Map

```{r prediction-map}

ggplot() +
  geom_sf(data = fishnet, aes(fill = predicted), color = NA) +
  scale_fill_viridis_c(option = "plasma") +
  labs(
    title = "Predicted Burglary Counts",
    fill = "Predicted"
  )
```


The predicted burglary map reveals clear spatial clustering of crime across 
Chicago. High predicted values are concentrated in specific hotspot areas, 
particularly in central and southern regions, indicating that the model is 
successfully identifying areas with elevated crime risk.

The model produces a smoother surface compared to raw crime counts, which is 
expected in regression-based predictions. This smoothing reflects the model’s 
ability to generalize underlying spatial patterns rather than overfitting to 
extreme observations.

The alignment between predicted hotspots and known high-crime areas suggests 
that the model captures key spatial dynamics effectively. In particular, the 
strong influence of spatial lag contributes to the formation of these clustered 
prediction patterns.



## Exercise 4.3: Spatial Cross Validation

```{r spatial-cv}

districts <- unique(fishnet$District)

cv_results <- map_dfr(districts, function(d) {
  
  train <- fishnet %>% filter(District != d)
  test  <- fishnet %>% filter(District == d)
  
  model <- glm.nb(
    crime_count ~ garbage_count + crime_lag,
    data = train
  )
  
  test$pred <- predict(model, newdata = test, type = "response")
  
  tibble(
    district = d,
    rmse = sqrt(mean((test$crime_count - test$pred)^2, na.rm = TRUE))
  )
})

cv_results
```

```{r cv-summary}

mean(cv_results$rmse)
```

Spatial cross-validation using a leave-one-district-out (LOGO) approach produced 
an average RMSE of approximately **13.11 burglaries per grid cell**.

The RMSE values vary substantially across districts, ranging from as low as 
approximately **2.3** to over **21**, indicating uneven predictive performance 
across the city. This variation suggests that the model performs well in some 
districts but struggles in others, particularly in areas with more complex or 
volatile crime patterns.

The relatively higher RMSE values compared to what would be expected under random 
cross-validation highlight the challenge of spatial generalization. When the model 
is tested on entirely unseen geographic areas, prediction errors increase, 
demonstrating that crime patterns are highly localized.

Overall, these results emphasize that while the model captures general spatial 
trends, its ability to transfer across different neighborhoods is limited, 
reflecting underlying spatial heterogeneity in urban crime dynamics.




## Conclusions

This analysis demonstrates that burglary patterns in Chicago are strongly shaped 
by spatial dynamics. The results show that crime is not randomly distributed but 
highly clustered, with nearby areas exerting significant influence on each other.

The Negative Binomial model confirms that **spatial lag is the most important 
predictor**, highlighting the importance of spatial dependence in crime modeling. 
Garbage cart density is also statistically significant, suggesting that it 
captures underlying neighborhood characteristics such as density or activity 
levels.

The model achieves moderate predictive performance, with an average RMSE of 
approximately **13 burglaries per grid cell** under spatial cross-validation. 
However, the variation in RMSE across districts indicates that prediction accuracy 
is uneven and context-dependent.

These findings underscore two key insights:  
1. **Incorporating spatial features is essential for accurate crime prediction**  
2. **Models struggle to generalize across different neighborhoods due to spatial heterogeneity**

From a policy perspective, this suggests that city interventions should be 
**place-specific rather than one-size-fits-all**, and predictive models should be 
used alongside local knowledge when allocating resources.

Overall, the analysis demonstrates the value of combining spatial data, 
neighborhood indicators, and appropriate modeling techniques to better understand 
and predict urban crime patterns.





